\tit

### 2021-02-13 paper.tex
wording

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the prefix ``{◬.}'' (module-specific names).
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, we reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type.
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the prefix ``{◬.}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\agda{api}}.

Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}}.
These definitions are meant to be read in ``row-major'' order, i.e., each row is one definition.
The \AF{Lang} representation is coinductive, and definitions are by \emph{copattern} matching \citep{AbelPientka2016}.
Note that the definitions are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
(The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)

The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
Fortunately, this issue was already solved by Andreas \citet{Abel2016}, also in the setting of trie-based language recognition, by using sized types \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\agda{api}}.

Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), but this time the compiler successfully proves \emph{total} correctness (including termination).

>>>>

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\agda{api}}.

Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}}.
These definitions are meant to be read in ``row-major'' order, i.e., each row is one definition.
Note that the definitions are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
(The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)

The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\agda{api}}.

Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), and this time the compiler successfully proves \emph{total} correctness (including termination).
